HAMdetector: A Bayesian regression model that integrates information to detect HLA-associated mutations

### 2021-03-25 docs/main.tex
a few small corrections and remarks

Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise. These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse. This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, with the confounding factors HLA linkage disequilibrium and codon covariation, e.g. in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.

@@

\subsection{Run times and convergence}
For a standard officie computer, run times of HAMdetector on the smaller HDV dataset were of the order of minutes and on the order of hours for the Hepatitis B dataset. For the large HIV dataset, the models were run overnight. Run times scale approximately linearly with the product \(NK\), where \(N\) is the number of sequences and \(K\) is the number of replacements. All model fits showed no signs of inference issues. In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 300 warm-up iterations. The effective sample size exceeded 500 for all model parameters, $\hat{R}$ convergence diagnostic values were below 1.01 in all cases.

@@

\begin{table}[h!]
  \caption{List of HAMs predicted Fisher's exact test (FET). In the last column a ``+'' marks experimental confirmation and a ``-'' a failed confirmation. Column ``post.prob.'' shows the corresponding posterior probabilities for positive associations computed with HAMdetector.}
  \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & ? \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}

>>>>

Sparsity-promoting priors \citep{Piironen2017} can drastically improve predictive performance, because the model is better able to differentiate between signal and noise. These priors convey the a priori expectation that most coefficients in a regression model are close to 0, i.e.\ that non-zero coefficients are sparse. This assumption is likely correct for HLA-associated mutations, because the number of epitopes that are restricted by a given HLA allele is typically small compared to the number of all possible epitopes. As presentation on MHC I is a necessary condition for the HLA system to function, most associations between an HLA allele and a replacement are likely truly zero, though sometimes confounding factors such as HLA linkage disequilibrium may lead to the appearance of sizeable associations.
\hl{DaHo: kann man so nicht schreiben, weil die kompensatorischen Mutationen zu *echten* Assoziationen führen: ``and codon covariation,
e.g. in the form of compensatory mutations that lie outside an epitope~\citep{Ruhl2011}.'' Ich finde trotzdem den Punkt der kompensatorischen Mutationen wichtig; sollte vielleicht in der Diskussion aufgegriffen werden. Z.B. ist es interessant zu sehen, ob unser Verfahren bei Positionen außerhalb von Epitopen immer noch einen Vorteil bei der Erkennung von Assoziationen liefert, oder zumindest solche Assoziationen wie in \citet{Ruhl2011} noch erkennt.}

@@

\subsection{Run times and convergence}
For a standard office computer, run times of HAMdetector on the smaller HDV dataset were of the order of minutes and on the order of hours for the Hepatitis B dataset. For the large HIV dataset, the models were run overnight. Run times scale approximately linearly with the product \(NK\), where \(N\) is the number of sequences and \(K\) is the number of replacements. All model fits showed no signs of inference issues. In total, samples were drawn from four Hamiltonian Markov chains with 1000 iterations each after 300 warm-up iterations. The effective sample size exceeded 500 for all model parameters, $\hat{R}$ convergence diagnostic values were below 1.01 in all cases.

@@

\begin{table}[h!]
  \caption{List of HAMs predicted by Fisher's exact test (FET). In the last column ``+'' and ``-'' mark successful and failed experimental confirmation, respectively. ``?'' are untested bona fide predictions. ``post.prob.'' are posterior probabilities for positive associations computed with HAMdetector.}
  \vspace{0.5cm}
  \begin{tabular}{c|c|c|c|c}
  replacement & allele  & p-value (FET) & post.~prob. & confirmed \\
  \hline
  S170N       & B*15 & $3\cdot10^{-8}$      & 0.99                                & +                        \\
  D101E       & B*37 & 0.0002                        & 0.96                                & +                        \\
  R105K       & B*27 & 0.0011                        & 0.93                                & +                        \\
  R139K       & B*41 & 0.0034                        & 0.92                                & +                        \\
  E47D        & B*18 & 0.0027                        & 0.90                                & +                        \\
  D33E        & B*13 & 0.0001                        & 0.86                                & -                        \\
  T134A       & A*68 & 0.0045                        & 0.82                                & -                        \\
  K43R        & B*13 & 0.0021                        & 0.77                                & -                        \\
  P89T        & B*37 & 0.0011                        & 0.75                                & +                        \\
  D47E        & A*30 & 0.0010                        & 0.76                                & -                        \\
  K113R       & B*13 & 0.0043                        & 0.76                                & -                        \\
  A107T       & B*14 & 0.0028                        & 0.70                                & -                        \\
  P49L        & A*30 & 0.0031                        & 0.63                                & -                        \\
  Q100L       & B*13 & 0.0018                        & 0.60                                & -                        \\
  D96E        & B*13 & 0.0035                        & 0.51                                & -  \\\hline
  E46D & A*02 & 0.0054 & 0.97 & ? \\
  V81I & A*68 & 0.0073 & 0.97 & ? \\
  K113N & B*08 & 0.0063 & 0.96 & ? \\
  A71T & B*41 & 0.0065 & 0.96 & ? \\
  L188I & A*68 & 0.0632 & 0.94 & ? \\
  T95S & A*01 & 0.0285 & 0.93 & ? \\
  D33E & A*03 & 0.0226 & 0.93 & ? \\
  P49L & B*13 & 0.0035 & 0.92 & ? \\
  A74S & A*68 & 0.0123 & 0.91 & ? \\
  E29D & B*44 & 0.0559 & 0.91 & ? \\
  D46E & B*57 & 0.0190 & 0.91 & ? \\
  R88K & A*68 & 0.0123 & 0.91 & ? \\
  T149P & B*52 & 0.0281 & 0.91 & ? \\
  K43R & A*02 & 0.2158 & 0.90 & ? \\
  N22S & B*08 & 0.0405 & 0.90 & ? \\
  \end{tabular}
  \label{tab:false-positives}
 \end{table}
